Carbon Nucleosynthesis

### 2023-04-28 thesis/tex/main.tex
bibliography updates; isolated v21 code

\newcommand{\citetjack}{Roberts et.\ al (2023, in prep.)}
\newcommand{\citepjack}{(Roberts et.\ al 2023, in prep.)}
\newcommand{\citealtjack}{Robert et.\ al 2023, in prep.}

@@

\begin{titlepage}
   \begin{center}
       \vspace*{5\baselineskip}
       \textbf{The Galactic Chemical Evolution of Carbon}\\
       Implications for Stellar Nucleosynthesis\\
       \vspace*{3\baselineskip}
        Undergraduate Research Thesis\\
       \vspace*{3\baselineskip}
    Presented in partial fulfillment of the requirements for graduation \textit{with research distinction in Astronomy and Astrophysics} in the College of Arts and Sciences of The Ohio State University\\
       \vspace*{3\baselineskip}
        by\\
       \vspace*{3\baselineskip}
       {Daniel Alexander Boyea}\\
       \vspace*{3\baselineskip}
       The Ohio State University\\
       April 2023\\
       \vspace*{3\baselineskip}
       Project Advisor: Professor David H. Weinberg, Department of Astronomy
       \vfill
   \end{center}
\end{titlepage}

@@

Galactic chemical evolution (GCE) aims to understand the chemical and star formation histories of galaxies. At the core of GCE models are stellar yields---the amount of each chemical element stars produce (or destroy). Each elements is produced in different amounts by different stars, leaving traces of the Galaxy's evolutionary history. Through \textit{Galactic Archeology}, we can reconstruct the history of our galaxy by investigating clues left behind in stars.

@@

To this end, I use a sample of subgiant stars from APOGEE \citep{apogee17}. According to stellar evolution models (CITES), these stars have not yet experienced FDU. Their phosphoric abundances should thus represent their birth composition.  In Fig.~\ref{fig:subgiants}, I plot this sample, selected by the criteria in Roberts et al. (2023, in prep; see Appendix \ref{sec:jack}). These empirical results contain information on the relative C and Mg yields, which I use later in this work.

@@

In Fig.~\ref{fig:y_cc}, I plot calculations of the IMF-integrated yields, defined with Eq.~\ref{eq:imf-yield} (computed using \VICE's \texttt{vice.yields.ccsne.fractional} function). 
CCSNe models predict a wide range of C yields, spanning almost 1 dex. 
Both the \citet{NKT13} and \cite{LC18} models show positive metallicity dependence. 
The \cite{LC18} models also include rotation, showing that variations in the rotational velocity of the star can dramatically increase C production and the metallicity dependence of $\Ycc$. Rotation induces more mixing allowing the CO core to grow larger. Our fiducial model's steep metallicity dependence near $Z\approx Z_\odot$ could be explained by rotation. 
Fig.~\ref{fig:ycc} shows the C11 model for comparison on the top. Especially at $Z\approx Z_\odot$, most CCSNe models dominate AGB C production. I will later also show empirically that CCSNe should dominate C production.

@@

CCSNe models also do not reach [O/Mg] due to overproduction of O, or underproduction of Mg, or both \citep{emily+21}. Here, I assume ${\rm [O/Mg]} = 0$, which is not compatible with CCSNe models but is consistent with APOGEE observations \citep{Weinberg+19, Weinberg+22}.

@@

As the strength of outflows controls the resulting $\alpha$ abundances, \JJ~create a metallicity gradient by defining
\begin{equation}
\eta(R) = \frac{y_{\alpha}^{\rm CC}}{Z_{\alpha, \odot}} 10^{(-0.08\text{ kpc}^{-1})(R-4\text{ kpc})+0.3} + r - 1.
\end{equation}
This choice of $\eta(R)$ results in a [$\alpha$/H] gradient consistent with Milky Way observations \citep{james+21}. I explore models with reduced outflows in \S\ref{sec:outflows}; however, the overall scaling of yields is unknown and this model is insensitive to this normalization.

@@

\citet{james+23} compare their model against the \cite{vincenzo+21} sample of APOGEE \citet{apogee17} red giants whose C and N abundance have been corrected for mixing process using MESA stellar evolution models. 
I instead use the \citetjack~sample of APOGEE subgiants. Subgiant stars have not undergone FDU, so their atmospheric C and N abundances are still reflective of their birth abundances and do not need mixing corrections. I additionally only compare the models to the low-$\alpha$ sequence.  The selection criteria and differences between the samples are described in Appendix \ref{sec:jack}

\section{The Evolution of C Across the Galaxy and Through Time}

Here, I present the time evolution of our fiducial model. In the next sections, I will discuss the choice of parameters and agreement with observations. Our fiducial model uses the C11 AGB yield tables (see \S\ref{sec:agb}, and Table \ref{tab:fiducial_mod}), and I amplify the C11 yields by a factor of 2.9 so that $f_{\rm AGB} = 0.2$.

The fiducial model has the following qualitative characteristics of its C yields: (a) C is predominantly produced in CCSNe, (b) CCSNe produce more C at higher metallicities, and (c) AGB stars produce less carbon at higher metallicities.

The evolution proceeds as follows.
Initially, CCSNe dominate production. As $\Ycc$ has strong metallicity dependence, [C/Mg] increases with time. Shortly thereafter, AGB stars contribute delayed C, increasing [C/Mg] even more steeply. As Mg begins to reach equilibrium, the [C/Mg] ratio plateaus as C approaches equilibrium. Finally, as $\Ycagb$ decreases or even becomes negative with higher metallicity, the [C/Mg] abundance may decline slightly.

I show the time evolution tracks of the fiducial model in Fig.~\ref{fig:c_evo} for \caah~ and \caafe~. Comparing [C/$\alpha$] against [Mg/Fe] enables us to see the late-time evolution of C more clearly. Because of the extended DTD of SNeIa enrichment, [Fe/H] takes longer to reach equilibrium as half of Fe production comes from type-Ia supernovae, and the late time evolution is not as clustered as [$\alpha$/H].

@@

Stellar measurements of Mg are more reliable, but O abundances are easier to measure in HII regions.  I therefore use Mg abundances from the \citetjack~subgiant sample as my primary empirical constraint. However, I use O abundances for gas-phase measurements and low-metallicity stars (\S \ref{sec:gas}). I model both Mg and O as metallicity-independent population-averaged yields dominated by massive stars. ${\rm [O/Mg]}\approx 0$ across a wide range of metallicities in APOGEE \citep{Weinberg+19, Weinberg+22}. 
I adopt O and Mg relative yields given by the solar mixture of \cite{asplund+09}.

>>>>

\newcommand{\citetjack}{Roberts et al.~(2023, in prep.)}
\newcommand{\citepjack}{(Roberts et al.~2023, in prep.)}
\newcommand{\citealtjack}{Robert et al.~2023, in prep.}

@@

\begin{titlepage}
   \begin{center}
       \vspace*{5\baselineskip}
       \textbf{The Galactic Chemical Evolution of Carbon}\\
       Implications for Stellar Nucleosynthesis\\
       \vspace*{3\baselineskip}
        Undergraduate Research Thesis\\
       \vspace*{3\baselineskip}
    Presented in partial fulfillment of the requirements for graduation \textit{with research distinction in Astronomy and Astrophysics} in the College of Arts and Sciences of The Ohio State University\\
       \vspace*{3\baselineskip}
        by\\
       \vspace*{3\baselineskip}
       {Daniel Alexander Boyea}\\
       \vspace*{3\baselineskip}
       The Ohio State University\\
       April 2023\\
       \vspace*{3\baselineskip}
       Project Advisor: Professor David H. Weinberg, Department of Astronomy \\
       James W. Johnson 
       \vfill
   \end{center}
\end{titlepage}

@@

Galactic chemical evolution (GCE) aims to understand the chemical and star formation histories of galaxies. At the core of GCE models are stellar yields---the amount of each chemical element stars produce (or destroy). Each element is produced in different amounts by different stars, leaving traces of the Galaxy's evolutionary history. Through \textit{Galactic Archeology}, we can reconstruct the history of our galaxy by investigating clues left behind in stars.

@@

To this end, I use a sample of subgiant stars from APOGEE \citep{apogee17}. According to stellar evolution theory and observations \citep{gilroy89, korn+07, lind+08, souto+18, souto19}, these stars have not yet experienced FDU but have well-mixed photospheres. Their photosphoric abundances should thus represent their birth composition.  In Fig.~\ref{fig:subgiants}, I plot this sample, selected by the criteria in Roberts et al. (2023, in prep; see Appendix \ref{sec:jack}). These empirical results contain information on the relative C and Mg yields, which I use later in this work.

@@

In Fig.~\ref{fig:y_cc}, I plot calculations of the IMF-integrated yields, defined with Eq.~\ref{eq:imf-yield} (computed using \VICE's \texttt{vice.yields.ccsne.fractional} function). 
CCSNe models predict a wide range of C yields, spanning almost 1 dex. 
Both the \citet{NKT13} and \cite{LC18} models show positive metallicity dependence. 
The \cite{LC18} models also include rotation, showing that variations in the rotational velocity of the star can dramatically increase C production and the metallicity dependence of $\Ycc$. Rotation induces more mixing allowing the CO core to grow larger. Our fiducial model's steep metallicity dependence near $Z\approx Z_\odot$ could be explained by rotation. 
Fig.~\ref{fig:y_cc} shows the C11 model for comparison on the top. Especially at $Z\approx Z_\odot$, most CCSNe models dominate AGB C production. I will later also show empirically that CCSNe should dominate C production.

@@

CCSNe models also do not reach [O/Mg] due to overproduction of O, or underproduction of Mg, or both \citep{emily+21}. Here, I assume ${\rm [O/Mg]} = 0$, which is not compatible with CCSNe models but is consistent with APOGEE observations \citep{weinberg+19, weinberg+22}.

@@

As the strength of outflows controls the resulting $\alpha$ abundances, \JJ~create a metallicity gradient by defining
\begin{equation}
\eta(R) = \frac{y_{\alpha}^{\rm CC}}{Z_{\alpha, \odot}} 10^{(-0.08\text{ kpc}^{-1})(R-4\text{ kpc})+0.3} + r - 1.
\end{equation}
This choice of $\eta(R)$ results in a [$\alpha$/H] gradient consistent with Milky Way observations \citep[e.g.][]{hayden+14, weinberg+19, frinchaboy+13}.

@@

\citet{james+23} compare their model against the \cite{vincenzo+21} sample of APOGEE \citep{apogee17} red giants whose C and N abundance have been corrected for mixing process using MESA stellar evolution models. 
I instead use the \citetjack~sample of APOGEE subgiants. Subgiant stars have not undergone FDU, so their atmospheric C and N abundances are still reflective of their birth abundances and do not need mixing corrections. I additionally only compare the models to the low-$\alpha$ sequence.  The selection criteria and differences between the samples are described in more detail in Appendix \ref{sec:jack}.

\section{The Evolution of C Abundances in the Galaxy}

Here, I present the time evolution of our fiducial model. In the next sections, I will discuss the choice of parameters and agreement with observations. 
The fiducial model has the following qualitative characteristics of its C yields: (a) C is predominantly produced in CCSNe, (b) CCSNe produce more C at higher metallicities, and (c) AGB stars produce less carbon at higher metallicities. Our fiducial model uses the C11 AGB yield tables (see \S\ref{sec:agb}, and Table \ref{tab:fiducial_mod}), and I amplify the C11 yields by a factor of 2.9 such that AGB stars account for 20\% of carbon at $Z=Z_\odot$.

I show the time evolution tracks of the fiducial model in Fig.~\ref{fig:c_evo} for \caah~ and \caafe~. Comparing [C/$\alpha$] against [Mg/Fe] enables us to see the late-time evolution of C more clearly. Because of the extended DTD of SNeIa enrichment, [Fe/H] takes longer to reach equilibrium as half of Fe production comes from type-Ia supernovae, and the late time evolution is not as clustered as [$\alpha$/H].
The evolution proceeds as follows.
Initially, CCSNe dominate production. As $\Ycc$ has strong metallicity dependence, [C/Mg] increases with time. Shortly thereafter, AGB stars contribute delayed C, increasing [C/Mg] even more steeply. As Mg begins to reach equilibrium, the [C/Mg] ratio plateaus as C approaches equilibrium. Finally, as $\Ycagb$ decreases or even becomes negative with higher metallicity, the [C/Mg] abundance may decline slightly.

@@

Stellar measurements of Mg are more reliable, but O abundances are easier to measure in HII regions.  I therefore use Mg abundances from the \citetjack~subgiant sample as my primary empirical constraint. However, I use O abundances for gas-phase measurements and low-metallicity stars (\S \ref{sec:gas}). I model both Mg and O as metallicity-independent population-averaged yields dominated by massive stars. ${\rm [O/Mg]}\approx 0$ across a wide range of metallicities in APOGEE \citep{weinberg+19, weinberg+22}. 
I adopt O and Mg relative yields given by the solar mixture of \cite{asplund+09}.
